Picky with peakpicking: assessing chromatographic peak quality with simple

### 2023-06-05 manuscript/manuscript.Rmd
Found the Kantz paper which does many of the same things

One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files. Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.[@Pirttila2022; @Guo2021; @Muller2020] Features that are high quality are typically represented in multiple files, especially in quality control pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely. Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert. Of course, this approach inherits all the difficulties of summary statistic calculations. The authors spent many hours discussing the implications of differences between mean and median values, but in general this approach produced more confidence than any single-file EIC.

@@

These two measures of success were then used to explore different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label. We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets. The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp. figure XX).

#### Performance relative to other peak picking models

```{r, eval=FALSE}
FT2040_features %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  with(table(feat_class, pred_class)) %>%
  caret::confusionMatrix(positive="Good", mode="everything")
# Falkor if everything was just classified as "Bad"
(980+56)/(980+56+45+196) # 0.824

@@

bind_rows(MS3000_features, FT2040_features) %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  with(table(feat_class, pred_class)) %>%
  caret::confusionMatrix(positive="Bad", mode="everything")
```

This model performs less well than previous reports in the literature, but is highly simplified. @Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5. However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so. This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis. This is perhaps a facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier.

@@

While the model we present here likely has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability. In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metric strength for later analyses.

@@

The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automatically and were implemented because they matched our intuition about what makes an MF look good to an MS expert. It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better. For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly. Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g. @Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective parameter. This metric could be improved with more careful summary statistics that account for the differences between samples. Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few. Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms.

@@

bind_rows(MS3000_features, FT2040_features) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_histogram(aes(x=mean_mz, fill=feat_class)) +
  facet_wrap(~feat_class, ncol = 1, scales = "free_y")
bind_rows(MS3000_features, FT2040_features) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=ifelse(feat_class=="Good", 1, 0)) %>%
  ggplot(aes(x=mean_mz, y=feat_class)) +
  geom_jitter(height=0.1, alpha=0.2) +
  geom_smooth(method="glm", method.args=list(family=binomial), formula = "y~x")
```

>>>>

One of the major ways in which this manuscript differs from prior work is its focus on summary statistics calculated across multiple files. Most existing peakpicking literature uses the single-file EIC peak as the core bit of training data, but that approach ignores critical information obtained elsewhere in the MS run that can change the judgement made on a single chromatogram.[@Pirttila2022; @Guo2021; @Muller2020] Features that are high quality are typically represented in multiple files, especially in quality control pooled samples, and a feature that only has a peak in a single file is typically regarded as highly suspect, if not removed entirely. Classifying an entire feature at once not only has the advantage of reducing the amount of manual labor by a factor equal to the number of files in the run (typically 10s-100s smaller) but is also a better representation of the judgement made by an MS expert. An exemplary implementation of this multi-file approach in prior work is reported in @Kantz2019, who compared the multi-file summary statistic model to a deep neural network and came to many similar conclusions.

@@

```{r, eval=FALSE}
both_min_model <- rbind(FT2040_features, MS3000_features) %>%
  select(feat_class, med_cor, med_SNR) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=feat_class=="Good") %>%
  glm(formula=feat_class~., family = binomial)

FT2040_features %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  with(table(feat_class, pred_class)) %>%
  caret::confusionMatrix(positive="Good", mode="everything")
# Falkor if everything was just classified as "Bad"
(980+56)/(980+56+45+196) # 0.824

@@

bind_rows(MS3000_features, FT2040_features) %>%
  mutate(pred_prob=predict(object=both_min_model,newdata = ., type = "response")) %>%
  mutate(pred_class=ifelse(pred_prob>0.5, "Good", "Bad")) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  with(table(feat_class, pred_class)) %>%
  caret::confusionMatrix(positive="Good", mode="prec_recall")
```

We explored several different types of classification models for separating good molecular features (MFs) from bad, with a particular focus on quantifying the likelihood of each class rather than just returning the label. We found that a simple two-parameter logistic regression model trained on two novel metrics of peak quality had reasonably good performance on the training set and was highly robust when applied to novel datasets. The logistic regression in particular was favored over the random forest and regularized regression we tried due to their similar performance and increased interpretability (Supp. figure XX).

This model outperforms the previously reported logistic regression model in @Kantz2019 and is highly simplified. There, they used a nineteen-parameter multiple logistic regression model and found a maximum performance of 80% GPF and an FDR of 34% on a cross-validated second cohort, similar to our cross-dataset testing. Our final two-parameter model also had an 80% GPF at a 0.5 likelihood threshold, but a significantly lower FDR of ~22.9%. This increased performance is likely due to our use of the metrics recalculated from the raw data, as their metrics were only calculated from the default MzMine2 peak parameters reported: similar to our XCMS-only model.

#### Performance relative to recent deep learning methods

@Guo2021 presented EVA and reported an accuracy of 90-95%, a range inclusive of our accuracy on both the Falkor (92.1%) and MESOSCOPE (94.4%) datasets when using a likelihood threshold of 0.5. However, we note that accuracy alone can be a highly misleading statistic to report when working with unbalanced datasets because very high accuracy can be obtained by simply classifying everything as bad, with a strong incentive to actually *increase* the number of bad peaks detected while doing so. This strategy, when applied to our data, returned accuracies in the 80-90% range despite being a useless classifier for downstream analysis. This is perhaps a facetious argument, but many of the single-parameter logistic models we tested never exceeded a 50% likelihood and essentially functioned as this type of classifier.

@@

While the model we present here likely has reduced performance relative to the CNNs, we would argue that its utility is not in maximizing performance but instead in maximizing interpretability, as previously noted by @Kantz2019. In particular, the CNNs provide no way to control the tradeoff between false positives and false negatives and no relative ranking of individual MF quality beyond the broad bins into which they are placed or explanation of relative metric strength for later analyses.

@@

The two metrics in the final model were derived from the raw EIC data rather than any summary statistic calculated by XCMS automatically and were implemented because they matched our intuition about what makes an MF look good to an MS expert. It is worth noting that these are very simple metrics and therefore fast to calculate, but we expect that more complicated metrics could perform even better. For example, the method of using the data within the peak boundaries for SNR calculation rather than data outside of them is not known to the authors to be implemented elsewhere but could be further improved by more advanced smoothing methods rather than using the residuals directly. Additionally, the calculation of peak shape using a Pearson's correlation to an idealized curve was not expected to be especially powerful given prior research (e.g. @Ipsen2010) and that the centWave algorithm essentially uses this information already during the wavelet fitting, but still proved to be a highly effective parameter. This metric could be improved with more careful summary statistics that account for the differences between samples. Currently, the use of the overall median value does a reasonable job at identifying MFs that appear in many samples but performs poorly when detecting MFs that appear in only a few. Also worth noting is that the calculation of any new metrics such as these that rely on access to the raw data require exact specification of the maximum and minimum *m/z* and retention time for a peak, values that are not always returned by peakpicking algorithms and must be recalculated, as in @Kantz2019.

@@

bind_rows(MS3000_features, FT2040_features) %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  ggplot() +
  geom_histogram(aes(x=mean_mz, fill=feat_class)) +
  facet_wrap(~feat_class, ncol = 1, scales = "free_y")
bind_rows(, FT2040_features)
MS3000_features %>%
  filter(feat_class%in%c("Good", "Bad")) %>%
  mutate(feat_class=ifelse(feat_class=="Good", 1, 0)) %>%
  ggplot(aes(x=mean_mz, y=feat_class)) +
  geom_jitter(height=0.1, alpha=0.2) +
  geom_smooth(method="glm", method.args=list(family=binomial), formula = "y~x")
```
